# Working with dates

One of the most frustrating things in data is working with dates. Everyone has a different opinion on how to record them, and every software package on the planet has to sort it out. Dealing with it can be a little ... confusing. And every dataset has something new to throw at you. So consider this an introduction. 

We're going to do this two ways. First I'm going to show you how to use base R to solve a tricky problem. And then we'll use a library called `lubridate` to solve a more common and less tricky problem. And then we'll use a new library to solve most of the common problems before they start. 

### The hard way

First, we'll import `dyplyr` like we always do. But we're going to start using a new library, which is part of the Tidyverse called ReadR. ReadR has some tools built in to solve some common import problems. 

```{r}
library(dplyr)
library(readr)
```

I'll give you an example of how ReadR is going to help later. We're going to use a dataset of [parking tickets at UNL](https://www.dropbox.com/s/gjfde8elm5hg3bd/tickets.csv?dl=0). If we do this the old way, this is what we get:

```{r}
tickets <- read.csv("../../Data/tickets.csv")
head(tickets)
```
Note the date is a factor. We have to fix that. There's a lot of ways to fix dates. The base R way is to use formatting. The code is ... a little odd ... but you have to strip the date into it's component parts then reassmble it into a date using formatting. 

```{r}
tickets %>% mutate(
    CleanDate = as.POSIXct(Date, format="%Y-%m-%d %H:%M:%S")
)
```

There is a [library called lubridate](https://cran.r-project.org/web/packages/lubridate/vignettes/lubridate.html) that can parse some common date problems. If it's not already installed, just run `install.packages('lubridate')`

```{r}
library(lubridate)
```

Lubridate can handle this tickets data easier with one of it's many functions. The functions parse dates given a basic pattern.

```{r}
tickets %>% mutate(
    CleanDate = ymd_hms(Date)
)
```
That's less code and less weirdness, so that's good. 

But to get clean data, I've installed a library and created a new field so I can now start to work with my dates. That seems like a lot, but don't think your data will always be perfect and you won't have to do these things. Still, there's got to be a better way. And there is. 

Fortunately, ReadR anticipates some date formattings and can automatically handle this. The change in your code? After adding the ReadR library, you just use `read_csv` instead of `read.csv`

```{r}
tickets <- read_csv("../../Data/tickets.csv")
head(tickets)
```

And just like that, the dates are formatted correctly. 

But you're not done with lubridate yet. It has some interesting pieces parts we'll use elsewhere. What's a question you might have about parking tickets on campus involving dates? How about what month are the most tickets issued? We could use formatting to create a Month field but that would group all the Aprils ever together. We could create a year and a month together, but that would give us an invalid date object and that would create problems later. Lubridate has something called a floor date that we can use. 

So to follow along here, we're going to use mutate to create a month field, group by to lump them together, summarize to count them up and arrange to order them. We're just chaining things together. 

```{r}
tickets %>% mutate(
    Month = floor_date(Date, "month")
) %>% 
  group_by(Month) %>% 
  summarise(
    total = n()
  ) %>%
  arrange(desc(total))
```

### Assignment 

ReadR doesn't always work, so it's useful to know how to apply it. Take [this data from the Lincoln Police Department](https://www.dropbox.com/s/3xxbr7h104zoenb/LPD_Dispatch_Records.csv?dl=0). Import it and view it with head.  

Note `RPT_Date`. You, as a human, can recognize it as a date. But even ReadR sees it as a number -- an integer. That won't do. So with Lubridate, create a new field with the proper date formatting and then using what you know, I want you to tell me what the 10 highest calls for service days are at the LPD since Jan. 1, 2017. How many of those were Husker home games?
How many calls for service happened on Saturday?

#### Rubric

1. Did you load the data correctly?
2. Did you apply the right date parsing formula?
3. Did you do the group and count correctly?
4. Did you do the sorts correctly?
5. Did you describe your steps in Markdown comments?