# Z scores

Z scores are a handy way to standardize scores so you can compare things across groupings. In our case, we may want to compare teams by year, or era. We can use z scores to answer questions like who was the greatest X of all time, because a Z score can put them in context to their era. 

We can also use z scores to ask how much better is team A from team B. 

So let's use Nebraska basketball, which if you haven't been reading lately is at a bit of a crossroads. 

A Z score is a measure of how far a number is from the population mean of that number. An easier way to say that -- how different is my grade from the average grade in the class. The formula for calculating a Z score is `(MyScore - AverageScore)/Standard Deviation of Scores`. The standard deviation is a number calculated to show the amount of variation in a set of data. In a normal distribution, 68 percent of all scores will be within 1 standard deviation, 95 percent will be within 2 and 99 within 3. 

## Calculating a Z score in R

```{r}
library(tidyverse)
```

Let's look at the current state of Nebraska basketball using the [same logs data we've been using, but for this season so far](). 

```{r}
gamelogs <- read_csv("data/logs20.csv")
```

The first thing we need to do is select some fields we think represent team quality:

```{r}
teamquality <- gamelogs %>% 
  select(Conference, Team, TeamFGPCT, TeamTotalRebounds, OpponentFGPCT, OpponentTotalRebounds)
```

And since we have individual game data, we need to collapse this into one record for each team. We do that with ... group by.

```{r}
teamtotals <- teamquality %>% 
  group_by(Conference, Team) %>% 
  summarise(
    FGAvg = mean(TeamFGPCT), 
    ReboundAvg = mean(TeamTotalRebounds), 
    OppFGAvg = mean(OpponentFGPCT),
    OffRebAvg = mean(OpponentTotalRebounds)
    )
```

To calculate a Z score in R, the easiest way is to use the scale function in base R. To use it, you use `scale(FieldName, center=TRUE, scale=TRUE)`. The center and scale indicate if you want to subtract from the mean and if you want to divide by the standard deviation, respectively. We do.

When we have multiple Z Scores, it's pretty standard practice to add them together into a composite score. That's what we're doing at the end here with `TotalZscore`. Note: We have to invert OppZscore and OppRebZScore by multiplying it by a negative 1 because the lower someone's opponent number is, the better. 

```{r}
teamzscore <- teamtotals %>% 
  mutate(
    FGzscore = as.numeric(scale(FGAvg, center = TRUE, scale = TRUE)),
    RebZscore = as.numeric(scale(ReboundAvg, center = TRUE, scale = TRUE)),
    OppZscore = as.numeric(scale(OppFGAvg, center = TRUE, scale = TRUE)) * -1,
    OppRebZScore = as.numeric(scale(OffRebAvg, center = TRUE, scale = TRUE)) * -1,
    TotalZscore = FGzscore + RebZscore + OppZscore + OppRebZScore
  )  
```

So now we have a dataframe called `teamzscore` that has 353 basketball teams with Z scores. What does it look like? 

```{r}
head(teamzscore)
```

A way to read this -- a team at zero is precisely average. The larger the positive number, the more exceptional they are. The larger the negative number, the more truly terrible they are. 

So who are the best teams in the country? 

```{r}
teamzscore %>% arrange(desc(TotalZscore))
```

Don't sleep on the Anteaters come tournament time!

But closer to home, how is Nebraska doing.

```{r}
teamzscore %>% 
  filter(Conference == "Big Ten") %>% 
  arrange(desc(TotalZscore))
```

So, as we can see, with our composite Z Score, Nebraska is ... not good. Not good at all. 