# Ridge charts

Ridgeplots are useful for when you want to show how different groupings compare with a large number of datapoints. So let's look at how we do this, and in the process, we learn about ggplot extensions. The extensions add functionality to ggplot, which doesn't out of the box have ridgeplots (sometimes called joyplots). 

In the console, run this: `install.packages("ggridges")`

Now we can add those libraries. 

```{r}
library(tidyverse)
library(ggridges)
```

So for this, let's look at every basketball game played since the 2014-15 season. That's more than 28,000 basketball games. [Download that data here](https://unl.box.com/s/ugd32b8aid883xjy5na6ak33ni7lix84).

```{r}
logs <- read_csv("~/Box/NCAABasketballData/logs1519.csv")
```

So I want to group teams by wins. Wins are the only thing that matter -- ask Tim Miles. So our data has a column called W_L that lists if the team won or lost. The problem is it doens't just say W or L. If the game went to overtime, it lists that. That complicates counting wins. And, with ridgeplots, I want to be be able to separate EVERY GAME by how many wins the team had over a SEASON. So I've got some work to do.

First, here's a trick to find a string of text and make that. It's called `grepl` and the basic syntax is grepl for this string in this field and then do what I tell you. In this case, we're going to create a new field called winloss look for W or L (and ignore any OT notation) and give wins a 1 and losses a 0. 


```{r}
winlosslogs <- logs %>% mutate(winloss = case_when(
  grepl("W", W_L) ~ 1, 
  grepl("L", W_L) ~ 0)
)
```

Now I'm going to add up all the winlosses for each team, which should give me the number of wins for each team. 

```{r}
winlosslogs %>% group_by(Team, Conference, season) %>% summarize(TeamWins = sum(winloss)) -> teamseasonwins
```

Now that I have season win totals, I can join that data back to my log data so each game has the total number of wins in each season. 

```{r}
winlosslogs %>% left_join(teamseasonwins, by=c("Team", "Conference", "season")) -> wintotallogs
```

Now I can use that same `case_when` logic to create some groupings. So I want to group teams together by how many wins they had over the season. For no good reason, I started with more than 25 wins, then did groups of 5 down to 10 wins. If you had fewer than 10 wins, God help your program. 

```{r}
wintotallogs %>% mutate(grouping = case_when(
  TeamWins > 25 ~ "More than 25 wins",
  TeamWins >= 20 & TeamWins <=25 ~ "20-25 wins",
  TeamWins >= 15 & TeamWins <=19 ~ "15-19 wins",
  TeamWins >= 10 & TeamWins <=14 ~ "10-14 wins",
  TeamWins < 10 ~ "Less than 10 wins")
) -> wintotalgroupinglogs
```

So my `wintotalgroupinglogs` table has each game, with a field that gives the total number of wins each team had that season and labeling each game with one of five groupings. I could use `dplyr` to do group_by on those five groups and find some things out about them, but ridgeplots do that visually.

Let's look at the differences in rebounds by those five groups. Do teams that win more than 25 games rebound better than teams that win fewer games? 

The answer might surprise you. 

```{r}
ggplot(wintotalgroupinglogs, aes(x = TeamTotalRebounds, y = grouping, fill = grouping)) +
  geom_density_ridges() +
  theme_ridges() + 
  theme(legend.position = "none")
```

Answer? Not really. Game to game, maybe. Over five seasons? The differences are indistiguishable. 

How about assists?

```{r}
ggplot(wintotalgroupinglogs, aes(x = TeamAssists, y = grouping, fill = grouping)) +
  geom_density_ridges() +
  theme_ridges() + 
  theme(legend.position = "none")
```

There's a little better, especially between top and bottom.

```{r}
ggplot(wintotalgroupinglogs, aes(x = Team3PPCT, y = grouping, fill = grouping)) +
  geom_density_ridges() +
  theme_ridges() + 
  theme(legend.position = "none")
```

If you've been paying attention this semester, you know what's coming next.

```{r}
ggplot(wintotalgroupinglogs, aes(x = TeamFGPCT, y = grouping, fill = grouping)) +
  geom_density_ridges() +
  theme_ridges() + 
  theme(legend.position = "none")
```

#### Assignment

You've been hired by Fred Hoiberg to tell him how to win the Big Ten. He's not impressed with that I came up with. So what you need to do is look for a *composite* measure that produces a meaningful ridgeplot. What that means is you're going to mutate `wintotalgroupinglogs` one more time. Is the differential between rebounding meaningful instead of just the total? Or assists? Or something else? Your call. Your goal is to produce a ridgeplot that tells The Mayor he needs to focus on doing X better than the opponent to win a Big Ten title.

##### Rubric

1. Did you mutate the data correctly?
2. Did you produce a ridgeplot?
3. Does it show a difference between top teams and the rest?




