# Tables

But not a table. A table with features. 

Sometimes, the best way to show your data is with a table. R has a neat package called `formattable` and you'll install it like anything else with `install.packages('formattable')`. 

So what does it do? Let's gather our libraries and [get some data](https://unl.box.com/s/g3eeuogx8bog72ig28enuakhpdlbn394). 

```{r}
library(tidyverse)
library(formattable)
```


```{r}
offense <- read_csv("data/offensechange.csv")
```


Let's ask this question: Which college football team saw the greatest improvement in yards per game last regular season? The simplest way to calculate that is by percent change. 

```{r}
changeTotalOffense <- offense %>%
  select(Name, Year, `Yards/G`) %>% 
  spread(Year, `Yards/G`) %>% 
  mutate(Change=(`2018`-`2017`)/`2017`) %>% 
  arrange(desc(Change)) %>% 
  top_n(20)
```
We've output tables to the screen a thousand times in this class with `head`, but formattable makes them look good with very little code. 

```{r}
formattable(changeTotalOffense)
```

So there you have it. Illinois improved the most. Because Nebraska gave them a quarterback, but I digress. First thing I don't like about formattable tables -- the right alignment. Let's fix that. 

```{r}
formattable(changeTotalOffense, align="l")
```

Next? I forgot to multiply by 100. No matter. Formattable can fix that for us. 

```{r}
formattable(
  changeTotalOffense, 
  align="l",
  list(
    `Change` = percent)
  )
```

Something else not great? I can't really see the magnitude of the 2018 column. A team could improve a lot, but still not gain that many yards (ahem, UTEP). Formattable has embeddable bar charts in the table. They look like this. 

```{r}
formattable(
  changeTotalOffense, 
  align="l",
  list(
    `2018` = color_bar("#FA614B"), 
    `Change` = percent)
  )
```
That gives me some more to mess with. 

One thing you can do is set the bar widths to make them relative to each other, using a different function called a `normalize_bar`.
```{r}
formattable(
  changeTotalOffense, 
  align="r",
  list(
    `2018` = normalize_bar("#FA614B"), 
    `2017` = normalize_bar("#FA614B"), 
    `Change` = percent)
  )
```

Note: bookdown is formatting this weird. Your numbers won't look like this.

Another way to deal with this -- color tiles. Change the rectangle that houses the data to a color indicating the intensity of it. Again, UTEP stands out.

```{r}
formattable(
  changeTotalOffense, 
  align="r",
  list(
     area(col = 2:3) ~ color_tile("#FFF6F4", "#FA614B"),
    `Change` = percent)
  )
```

### Exporting tables

The first thing you need to do is install some libraries -- do this in the console, not in an R Studio code block because htmltools get's a little weird. 

```
install.packages("htmltools")
install.packages("webshot")

webshot::install_phantomjs()
```

Now, copy, paste and run this code block entirely. Don't change anything. 

```{r}
library("htmltools")
library("webshot")    

export_formattable <- function(f, file, width = "100%", height = NULL, 
                               background = "white", delay = 0.2)
    {
      w <- as.htmlwidget(f, width = width, height = height)
      path <- html_print(w, background = background, viewer = NULL)
      url <- paste0("file:///", gsub("\\\\", "/", normalizePath(path)))
      webshot(url,
              file = file,
              selector = ".formattable_widget",
              delay = delay)
    }
```

Now, save your formattable table to an object using the `<-` assignment operator. 

After you've done that, you can call the function you ran in the previous block to export as a png file. In my case, I created an object called table, which is populated with my formattable table. Then, in export_formattable, I pass in that `table` object and give it a name. 

```{r}
table <- formattable(
  changeTotalOffense, 
  align="r",
  list(
     area(col = 2:3) ~ color_tile("#FFF6F4", "#FA614B"),
    `Change` = percent)
  )

export_formattable(table,"table.png")
```

For now, pngs are what you need to export. There is a way to export PDFs, but they lose all the formatting when you do that, which is kind of pointless. 